Carp\'{e} Data: Supporting Serendipitous Data Integration in Personal Information Management

### 2012-09-15 chi2012.tex
interaction description
minor edits

\begin{abstract}
Exploratory sequential mixed-method study
\end{abstract}

@@

In recent years, an unprecedented quantity and variety of information has been made available as structured data on the Web, through APIs, datasets, and data feeds.  This information includes: access to data previously hidden behind services and applications, such as retailer product catalogues; information not previously directly released to the public, such as open government data or records of personal financial transactions; and new kinds of data generated by emerging kinds of data sources, such as wearable and smartphone-based biosensors.  A primary goal of opening up such direct access to information is that end-user citizens will make more informed decisions pertaining to their personal information, for example health, wealth, and well-being \cite{}.  This data is predominately used by  specialised app developers, journalists and other ``data specialists'' but remains inaccessible to citizens because of the available tools.  For example, the most commonly used structured personal information management (PIM) tools either manage a small, fixed set of data types (such as digital calendaring tools and  to-do list managers) or provide little or no support for structured data at all, such as text editors and sketching/drawing tools \cite{}.

We hypothesise that end-users would benefit from this rich data with tools allowing them to effectively browse, use and consume heterogeneous structured data from diverse sources.  In this paper, we present our investigation of how extending PIM tools to effectively use emerging ecosystems of personal data impact information practice.  We used a three stage sequential exploratory mixed-method approach: first, we performed a qualitative analysis of a semi-structured interview pre-study to understand the types of tasks people performed using a plurality of  information sources, and the processes that they relied upon to perform such tasks.  Our results suggested that people preferred to rely upon multiple, diverse sources to singular, integrated ones for a number of reasons, including coverage, reliability, and seeking out more suitable alternatives.

Second, we conducted a quantitative, structural analysis of various popular personal data sources available today to identify potential barriers to effective unification of data from each.  Drawing upon definitions from the database integration and automatic schema matching literature, we characterised the types of heterogeneity exhibited across data sources in three popular domains: social networking, shopping, and dining, finding that terminological heterogeneity dominates the simpler data feeds (including social networking and restaurant recommendations), while structural issues pervade the more complex schemas of online retailers' product catalogues.

Third, we developed DataPalette, which focuses on the most common types of heterogeneity to enable serendipitous ``data mixing'' - a form of simple integration sufficient to allow people to easily and effectively combine and compare information sources without the need to write code.  It was principally designed based upon people's data integration needs and to support data heterogeneity observed in the previous studies.  We embarked on a user-centric design exercise towards an interface to facilitate simple, ``light-touch'' integration tasks of diverse, heterogeneous data sources.  We then performed a usability evaluation of DataPalette, which revealed that most users were comfortable with the interactive integration mechanisms we devised, and that this interface effectively improved people's ability to perform multifaceted decision making tasks using multiple sources.

@@

In this section, we touch on motivating work for DataPalette from three fields: peronal information management, data integration, and end-user mashups and toolkits.

The primary motivation for this work stems from field studies in personal information management, where letting users effectively consolidate and work with data from multiple sources has remained a challenge for over a decade. The problem of data fragmentation (\cite{Jones05towardsa}, for example, derives from an inability to effectively consolidate data across multiple applications, services and platforms (e.g., \cite{bergman}, \cite{boardmansasse}). In our own previously-conducted field study, (reference anonymised) we observed variations of manual ``coping strategies'' (or, if programmers, elaborate, custom-coded one-off solutions) to allow people to manage heterogeneous data from disparate sources -- be them online, offline, or other people.  Similar findings from other studies include Voida et al.'s observations of volunteer coordinators' ``homebrew databases'', manually-mainted information assemblages concocted to handle the widely varied and heterogeneous data collection requirements such groups needed to coordinate their activities \cite{Voida:2011:HDC:1978942.1979078}.
%% TODO MORE HERE

@@

\subsection{PS.1 - Understanding Data Diversity in Everyday Tasks}
We designed a semi-structured interview about tasks people performed on-line that required the use of multiple websites, and the processes people used to perform such tasks.  We analysed the data collected from this study using a thematic approach, as described below.

Our pre-study investigated the types of tasks people performed online that required heterogeneous data sources, and which tools they would use to organise a large social event involving 15 or more people.  We interviewed 8 participants, using a general interview guide approach based loosely on a set of questions (see Appendix ?).  This approach allowed both the participant and interviewer a degree of freedom in conversation and each interview could adapt to the participant's experiences, while collecting the same general areas of information.  Each interview was recorded and notes taken, the interviewer was trained so that they were knowledgeable about the importance of the study and to minimise bias.

@@

\subsubsection{Question 2: Large Social Event}
All of the participants would organise a large social event by talking to their friends, about their ideas, preferences and recommendations.  The tools they would use to acquire this information included face-to-face, phone calls, email, Skype, Facebook events.  The method chosen depended on the time scale required to organise an event, in general if there was a short time scale people would speak in person or on the phone.  Otherwise, people would set up a Facebook event page to discuss ideas with their friends.  If their friends were not on Facebook they would email them.  All but one participant said they would not use Doodle\footnote{Doodle - \emph{www.doodle.com}} to organise day and time of an event, because they felt that people did not fill in the form and it was more suit to organising work event not social.

@@

\subsection{Interaction with DataPalette}
In order to feel familiar to new users and focus our design efforts on integration-related aspects of the interface, we borrowed metaphors and interactions directly from MacOS Finder-style file managers.  Specifically, in the DataPalette workspace, shown in Figure \ref{fig:workspace}, the user can arbitrarily drag and drop entities from data sources (on the left) into windows that they create, resize, arrange and label. (Although the design could be generalised to supporting nested groups, our prototype only supported a single level of grouping.)

Unlike file managers, the basic unit of data is not a file, but an \emph{instance} - corresponding to an RDF instance or any small bit of structured data. In order to make working with relational graphs manageable, an early decision we made was to break up RDF-like relational data) into discrete instances with properties; for JSON or XML data feeds, we represented all subtrees with more than 1 child as its an instance.

\subsubsection{Multipath selection: link-sliding for heterogeneous sets}
When working with instances, users may be interested in a particular aspect of an instance rather than just the instance itself; for example, a user may be interested in the price of a particular product, its rating, manufacturer, and so on.  These are represented as \emph{properties} of the item, listed under the name of the instance.  Selecting any of such properties creates a \emph{path   selection} which dereferences all of the instances in that group that have that property, yielding their corresponding values.  For example, selecting the \emph{manufacturer} property for an instance in a group will cause the manufacturer of that item to appear alongside its label, and will also cause the manufacturers of \emph{all other   instances in that group that have a manufacturer property} to be dereferenced (and displayed) as well.  When such a path selection is applied, the properties displayed are updated to be the set of properties of the \emph{terminating value of the path}, so that such value can be further dereferenced.  For example, once selecting a manufacturer, a user may want to know the product's manufacturer's overall reputation, so might select its ``reputation score'' property, thereby extending the path selection.

Instances that cannot be dereferenced by a particular path selection  are left alone;  a second path selection can be created for such insances, which, like the first path selection, causes all like instances to be dereferenced.  Each group can carry an arbitrary number of path selections, and each path selection can be extended arbitrarily deep by selecting successive properties of values. This achieves the ability to \emph{link slide} in the same way introduced by Parallax \cite{parallax}.

Each instance is displayed with its corresponding properties, and \emph{value selection}.

Data sources were represented as collapsable panels in a sidebar on the left of the workspace, from which instances or entire data sources could be ``pulled off'' onto the workspace.  Figure \ref{fig:workspace} displays the entire UI overview.

\subsection{Summary of interaction features}

\begin{itemize}
\item \emph{Group multi-path value selection} - Selecting any of the properties of an instance in a group (window) causes all of the rest of the instances that have that property in that group to be dereferenced along the same property, displaying the resulting value underneath the instance label.  Selecting a further property of the resulting value further dereferences this value.
\item \emph{Drag-and-drop coreference / value combination} - In order to specify that two instances correspond to the same entity, they can be simly dragged together
\item \emph{Group-based Drag-and-drop matching} - In order to specify that two instances correspond to the same entity, they can be simly dragged together
\item \emph{Universal model-based brushing} -
\end{itemize}

>>>>

\begin{abstract}
Exploratory sequential mixed method study
\end{abstract}

@@

In recent years, an unprecedented quantity and variety of information has been made available as structured data on the Web, through APIs, datasets, and data feeds.  This information includes: access to data previously hidden behind services and applications, such as retailer product catalogues; information not previously directly released to the public, such as open government data or records of personal financial transactions; and new kinds of data generated by emerging kinds of data sources, such as wearable and smartphone-based biosensors.  While a primary goal of opening up such direct access to information is that end-user citizens will make more informed decisions pertaining to their personal information, for example health, wealth, and well-being \cite{}.  This data is predominately used by  specialised app developers, journalists and other ``data specialists'' but remains inaccessible to citizens because of the available tools.  For example, the most commonly used structured personal information management (PIM) tools either manage a small, fixed set of data types (such as digital calendaring tools and  to-do list managers) or provide little or no support for structured data at all, such as text editors and sketching/drawing tools \cite{}.

We hypothesise that end-users would benefit from this rich data with tools allowing them to effectively browse, use and consume heterogeneous structured data from diverse sources.  In this paper, we aim to investigate how extending PIM tools to effectively use emerging ecosystems of personal data would impact information practice.  We use a three stage sequential exploratory mixed method approach: First, we performed a qualitative analysis, using a small semi-structured interview study to examine the types of tasks people perform requiring more than one source of information, and the processes that they would use to perform such a task.  Our results suggest that people prefer to rely upon multiple, diverse sources to singular integrated ones for a number of reasons, including coverage, reliability, and seeking out more suitable alternatives.

Second, we conducted a quantitative analysis of various popular personal data sources available today to identify potential barriers to effective unification of data from each.  We then draw upon definitions and work from the database integration systems literature to characterise types of heterogeneity exhibited across data sources in three popular domains: social networking, shopping, and dining, finding that terminological heterogeneity dominates the simpler data feeds (including social networking and restaurant recommendations), while structural issues pervade the more complex schemas of online retailers' product catalogues.

Third, we developed DataPalette, which focuses on the most common types of heterogeneity to enable serendipitous ``data mixing'' - a form of simple integration sufficient to allow people to easily and effectively combine and compare information sources without the need to write code.  It was principally designed based upon people's data integration needs and to support data heterogeneity observed in the previous studies.  We embarked on a user-centric design exercise towards an interface to facilitate simple, ``light-touch'' integration tasks of diverse, heterogeneous data sources.  We then performed a usability evaluation of DataPalette, which revealed that most users were comfortable with and that this interface is effectively expedites the most sophisticated of the typical kinds of ``data integration'' tasks people need to perform while planning personal.

@@

In this section, we touch on motivating work for DataPalette from three fields: peronsal information management, data integration, and end-user mashups and toolkits.

The primary motivation for this work stems from field studies in personal information management, where letting users effectively consolidate and work with data from multiple sources has remained a challenge for over a decade. The problem of data fragmentation (\cite{Jones05towardsa}, for example, derives from an inability to effectively consolidate data across multiple applications, services and platforms (e.g., \cite{bergman}, \cite{boardmansasse}). In our own previously-conducted field study, (reference anonymised) we observed variations of manual ``coping strategies'' (or, if programmers, elaborate, custom-coded one-off solutions) to allow people to manage heterogeneous data from disparate sources -- be them online, offline, or other people.  Similar findings from other studies include Voida et al.'s observations of volunteer coordinators' ``homebrew databases'', manually-maintained information assemblages concocted to handle the widely varied and heterogeneous data collection requirements such groups needed to coordinate their activities \cite{Voida:2011:HDC:1978942.1979078}.
%% TODO MORE HERE

@@

\subsection{PS.1 - Understanding Data Diversity in Everyday Tasks}
We designed a semi-structured interview about tasks people performed online that required the use of multiple websites, and the processes people used to perform such tasks.  We analysed the data collected from this study using a thematic approach, as described below.

Our pre-study investigated the types of tasks people performed online that required heterogeneous data sources, and which tools they would use to organise a large social event involving 15 or more people.  We interviewed 8 participants, using a general interview guide approach based loosely on a set of questions.  This approach allowed both the participant and interviewer a degree of freedom in conversation and each interview could adapt to the participant's experiences, while collecting the same general areas of information.  Each interview was recorded and notes taken, the interviewer was trained so that they were knowledgeable about the importance of the study and to minimise bias.

@@

\subsubsection{Question 2: Process Used to Organise a Large Social Event}
All of the participants would organise a large social event by talking to their friends, about their ideas, preferences and recommendations.  The tools they would use to acquire this information included face-to-face, phone calls, email, Skype, Facebook events.  The method chosen depended on the time scale required to organise an event, in general if there was a short time scale people would speak in person or on the phone.  Otherwise, people would set up a Facebook event page to discuss ideas with their friends.  If their friends were not on Facebook they would email them.  All but one participant said they would not use Doodle\footnote{Doodle - \emph{www.doodle.com}} to organise day and time of an event, because they felt that people did not fill in the form and it was more suit to organising work event not social.

@@

\subsection{Interaction mechanisms for data mixing}
In order to familiar to new users, we borrowed the overall interaction and feel from MacOS Finder-style file managers, with support for separate windows which can be freely moved, resized and titled. (As described in results, we discovered that windows caused considerable clutter at the later stages of some tasks, a fact we did not detect until the main study.)  Instead of files, however, the element of manipulation was a single, discrete structured data item, or \emph{instance}.  In order to define \emph{instances} for non-RDF data, such as JSON or XML, we defined all subtrees with more than 1 child as its an instance.  Data sources were represented as collapsable panels in a sidebar on the left of the workspace, from which instances or entire data sources could be ``pulled off'' onto the workspace.  Figure \ref{fig:workspace} displays the entire UI overview.

\begin{itemize}
\item \emph{Group multi-path value selection} - Selecting any of the properties of an instance causes all of the rest of the instances in that box to be dereferenced along the same property, displaying the resulting value underneath the instance label.  Selecting a further property of the resulting value further dereferences this value.
\item \emph{Drag-and-drop coreference / value combination} - In order to specify that two instances correspond to the same entity, they can be simly dragged together
\item \emph{Group-based Drag-and-drop matching} - In order to specify that two instances correspond to the same entity, they can be simly dragged together
\item \emph{Universal model-based brushing} -
\end{itemize}
